### 2020-01-18 nlp_models.tex
fixed GPT2 spectral norm
udpating nlp plots and tables

\charles{more here ?}
We analyze GPT models deployed with the popular HuggingFace PyTorch library.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ Layer Weight Matrices $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large, typically sparse weight matrices, and they can be
 \emph{poorly correlated} when trained with insufficient data--as we shall see below.

@@

The additional OpenAI GPT2 (English) models are: \emph{gpt-medium, got-large, and gpt-xl}, 
having include $12, 24, 36, and 48$ layers, resp., with increasingly larger weight matrices.
The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.
Table \ref{table:nlp} reports results for the average log norm metrics, using \emph{weightwatcher (0.2.7)},
and with fully reproducible Jupyter notebooks.\footnote{\url{https://github.com/CalculatedContent/kdd2020}}

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
\hline
   &    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
 Series & \#Layers   & $\Vert\mathbf{W}\Vert_{F}$ & $\Vert\mathbf{W}\Vert_{\infty}$ & $\hat{\alpha}=\alpha\log\lambda_{max}$ & $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$ \\
\hline
 GPT & & & & &\\
GPT2 & 50 & 2.05 &2.59& 9.78 & 10.03 \\
GPT2 medium & 98 & 2.08 &2.58& 9.74 & 10.01 \\
GPT2 large & 146 & 1.85 &1.99& 7.67 & 7.94 \\
GPT2 xl & 194 & 1.86 &1.92& 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average Log Norm Metrics for pretrainnd OpenAI GPT and GPT2 models.}
\label{table:nlp}
\end{table}

We find that in the pretrained GPT models, the first embedding layer(s) have unusually large eigenvalues,
reflecting a lack of normalization.  For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$, 
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier); including them does
not significantly alter the results here.

@@

\begin{figure}
    \centering
    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha.hist.png}
       }
    \qquad
    \subfigure[Log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm.hist.png}
       }

@@

    \subfigure[ Power Law Exponent $\alpha$  ]{
        \includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[ Spectral Norm $\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \label{fig:resnet-snorm-layer}
    }
    \qquad
    \subfigure[ Alpha-Norm $\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

For this series of GPT2 models, the average $\alpha$ still decreases with increasing model size,
althogh, the differences are less noticible than between the GPT and GPT2 models.
Unlike GPT, however, the Layer Spectral Norms $\vert\mathbf{W}\Vert_{\infty}$ 
and Alpha-Norms $\vert\mathbf{W}\Vert_{\alpha}^{\alpha}$
 behave  as expected for GPT2 layers, with the larger models 
consistently having  smaller norms.

@@

    \subfigure[ Power Law Exponent $\alpha$  ]{
        \includegraphics[width=4.5cm]{img/GPT2_all_alpha_hist.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[ Spectral Norm $\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT2_all_spectralnorm_hist.png}
        \label{fig:resnet-snorm-layer}
    }
    \qquad
    \subfigure[ Alpha-Norm $\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT2_all_alpha_weighted_hist.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Power Law Exponnents, Spectral Norm, and Alpha-Norm for different size models in the GPT2 architecture series.  Spectral Norm dhistogram does not contain the first 2 layers because the values are unusually large.}
    \label{fig:gpt-alpha-layers}
\end{figure}

>>>>

\charles{more here ?}
We analyze GPT models deployed with the popular HuggingFace PyTorch library.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ Layer Weight Matrices $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.

@@

The OpenAI GPT2 (English) models are: \nred{gpt-small, gpt-medium, gpt-large, and gpt-xl}, \
having include $12, 24, 36, \text{and }48$ layers, resp., with increasingly larger weight matrices.
The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.
Table \ref{table:nlp} reports results for the average log norm metrics, using \emph{weightwatcher (0.2.7)},
and with fully reproducible Jupyter notebooks.\footnote{\url{https://github.com/CalculatedContent/kdd2020}}

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
\hline
   &    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
 Series & \#Layers   & $\Vert\mathbf{W}\Vert_{F}$ & $\Vert\mathbf{W}\Vert_{\infty}$ & $\hat{\alpha}=\alpha\log\lambda_{max}$ & $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT (small) & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
GPT2 medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2 large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2 xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average Log Norm Metrics for pretrainnd OpenAI GPT and GPT2 models.}
\label{table:nlp}
\end{table}

@@

\begin{figure}
    \centering
    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha.hist.png}
       }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm.hist.png}
       }

@@

    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \label{fig:resnet-snorm-layer}
    }
    \qquad
    \subfigure[log Alpha-Norm $\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

Notice the first embedding layer(s) have unusually large Spectral Norms,
reflecting a lack of normalization.  For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier); including them does
not significantly alter the results here.

@@

For this series of GPT2 models, the average $\alpha$ still decreases with increasing model size,
althogh, the differences are less noticible than between the GPT and GPT2 models.
Unlike GPT, however, the Layer (log) Spectral Norms $\log\vert\mathbf{W}\Vert_{\infty}$ 
and (log) Alpha-Norms $\log\vert\mathbf{W}\Vert_{\alpha}^{\alpha}$
 behave  as expected for GPT2 layers, with the larger models  consistently having  smaller norms.

@@

    \subfigure[Power Law Exponent $\alpha$  ]{
        \includegraphics[width=4.5cm]{img/GPT2_all_alpha_hist.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT2*_all_spectralnorm_hist.png}
        \label{fig:resnet-snorm-layer}
    }
    \qquad
    \subfigure[log Alpha-Norm $\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT2_all_logpnorm_hist.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Power Law Exponnents, Spectral Norm, and Alpha-Norm for different size models in the GPT2 architecture series.  *Note: the log spectral norms (b) histogram omits the first 2 layers, corresponding to the embedding layer, which are normalized differently and have anamolously large values.}
    \label{fig:gpt-alpha-layers}
\end{figure}
